---
title: "Leucegene data analysis"
#author: "Farnoosh"
#date: '2023-03-16'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(survival)
library(survminer)
library(Rcpp)
library(PharmacoGx)
library(ggplot2)
library(qs)
library(meta)
library(forestplot)
library(readxl)

```

# Data

Consider the Leucegene AML study including 452 samples across GSE49642, GSE52656, GSE62190, GSE66917, GSE67039, and GSE106272 cohorts. 320 samples remain with available survival outcome (overall survival or OS) where 263 of them with RNA-seq expression values (i.e., log2(TPM + 1)) for the signature analysis including 9 (plus three more signatures) RNA signatures. Due to the lack of clinical data, cohort GSE106272 will be removed from the analyses and 263 samples with remain for the rest of analysis. Genes with zero variance also removed.  


```{r AML expression data}

dat_GSE49642 <- qread("~/Data/GSE49642_SE.qs")
dim(dat_GSE49642)

dat_GSE52656 <- qread("~/Data/GSE52656_SE.qs")
dim(dat_GSE52656)

dat_GSE62190 <- qread("~/Data/GSE62190_SE.qs")
dim(dat_GSE62190)

dat_GSE66917 <- qread("~/Data/GSE66917_SE.qs")
dim(dat_GSE66917)

dat_GSE67039 <- qread("~/Data/GSE67039_SE.qs")
dim(dat_GSE67039)

dat_GSE106272 <- qread("~/Data/GSE106272_SE.qs")
dim(dat_GSE106272)

dat <- list(GSE49642 = dat_GSE49642, 
            GSE52656 = dat_GSE52656, 
            GSE62190 = dat_GSE62190,
            GSE66917 = dat_GSE66917, 
            GSE67039 = dat_GSE67039, 
            GSE106272 = dat_GSE106272)

## extract expression data
dat_expr <- lapply(1:length(dat), function(k){
  
  as.data.frame(assay(dat[[k]]))
  
})

names(dat_expr) <- names(dat)

## sample annotation
dat_sample_annot <- lapply(1:length(dat), function(k){
  
   as.data.frame(colData(dat[[k]]))
  
})

names(dat_sample_annot) <- names(dat)

## fix sample names using expression data

dat_expr <- lapply(1:length(dat_expr), function(k){
  
  colnames(dat_expr[[k]]) <- dat_sample_annot[[k]]$title
  dat_expr[[k]]
  
})

names(dat_expr) <- names(dat)

## read clinical data (it is private and won't be curated and shared on ORCESTRA)

dat_clinical <- read_xlsx("~/Data/Leucegene.xlsx")
dat_clinical <- as.data.frame(dat_clinical)

## merge clinical and expression data
dat_expr_clinical <- lapply(1:(length(dat_expr)-1), function(k){
  
  int <- intersect(colnames(dat_expr[[k]]), dat_clinical$`ID BCLQ`)
  dat_rpkm <- dat_expr[[k]][ , int ]
  
## convert RPKM to TPM
dat_tpm <-  apply(dat_rpkm, 2, function(x) x / sum(as.numeric(x), na.rm = TRUE) * 10^6) %>% as.data.frame()
  dat_logTPM <- log2(dat_tpm+1)
  
  list(expr = dat_logTPM,
       clinical = dat_clinical[dat_clinical$`ID BCLQ` %in% int, ])
  
})

names(dat_expr_clinical)  <- names(dat)[-6]

# Integrate all cohorts

int <- intersect(intersect(intersect(intersect(rownames(dat_expr_clinical[[1]]$expr),            rownames(dat_expr_clinical[[2]]$expr)),
                 rownames(dat_expr_clinical[[3]]$expr)),
                 rownames(dat_expr_clinical[[4]]$expr)),
                 rownames(dat_expr_clinical[[5]]$expr)) # 19837


for(i in 1:length(dat_expr_clinical)){
  
  dat <- dat_expr_clinical[[i]]$expr
  dat_expr_clinical[[i]]$expr <- dat[rownames(dat) %in% int, ]

}


## merge expression data
dat_expr <-  dat_expr_clinical[[1]]$expr
for(i in 2:length(dat_expr_clinical)){
  
  dat_expr <- cbind(dat_expr, dat_expr_clinical[[i]]$expr)

}
  
get_zero_variance <- apply(dat_expr, 1, var)
rem <- which(get_zero_variance == 0)
dat_expr <- dat_expr[-rem, ]

## merge clinical data
dat_clinical <-  dat_expr_clinical[[1]]$clinical
for(i in 2:length(dat_expr_clinical)){
  
  dat_clinical <- rbind(dat_clinical, dat_expr_clinical[[i]]$clinical)

}
  

# signature data
dat_signature <- read.csv("~/Data/signature_updated_April.csv")
colnames(dat_signature)[c(2,12)] <- c("8gene", "8Drivers")

signature <- lapply(1:ncol(dat_signature ), function(k){
  
  sig_val <- dat_signature[, k]
  sig_val[sig_val != ""]

  })

names(signature) <- colnames(dat_signature)
str(signature)

```

## Batch Effect 

To assess any potential batches across cohorts after integration, the principal component analysis (PCA) was applied. The scatter plot shows no considerable variation across cohorts. 

```{r batch effect}

group <- c(rep("GSE49642", nrow(dat_expr_clinical$GSE49642$clinical)),
           rep("GSE52656", nrow(dat_expr_clinical$GSE52656$clinical)),
           rep("GSE62190", nrow(dat_expr_clinical$GSE62190$clinical)),
           rep("GSE66917", nrow(dat_expr_clinical$GSE66917$clinical)),
           rep("GSE67039", nrow(dat_expr_clinical$GSE67039$clinical)))

group <- factor(group)
cols <- c("#1a9850", "#9970ab", "#4575b4", "#762a83", "#8c510a")

pca_results <- prcomp(t(dat_expr),
             center = TRUE,
             scale = TRUE)

propve <- pca_results$sdev^2 / sum(pca_results$sdev^2)
barplot(round(propve*100, 2)[1:50], ylab = "Proportion of Variance Explained (%)",
                                    xlab = "principal component (top 50 PCs)")


plot(pca_results$x[,1],pca_results$x[,2],pch="+", xlab='PC1 (15.33%)',ylab='PC2 (8.04%)', 
     main='GEO AML Leucegene cohorts')
ind= group == levels(group)[1]
points(pca_results$x[ind,1],pca_results$x[ind,2], pch="+",col=cols[1])
ind= group == levels(group)[2]
points(pca_results$x[ind,1],pca_results$x[ind,2], pch="+",col=cols[2])
ind= group == levels(group)[3]
points(pca_results$x[ind,1],pca_results$x[ind,2], pch="+",col=cols[3])
ind= group == levels(group)[4]
points(pca_results$x[ind,1],pca_results$x[ind,2], pch="+",col=cols[4])
ind= group == levels(group)[5]
points(pca_results$x[ind,1],pca_results$x[ind,2], pch="+",col=cols[5])
legend('bottomleft',paste(levels(group)), pch="+",col=cols, cex=0.5)


```

## Signature score and survival analysis

To keep consistency with previous results using GEPIA2 web application, the signature score is computed using mean approach. The Cox regression model is considered to assess the association of signatures with survival outcome (OS) per cohort. Multiple test correction approach; Benjamini \& Hochberg; is applied to correct multiple tests by controlling the false discovery rate (FDR). 

To get the two groups High vs. Low for signature score, two type of thresholds were applied, median or 50 percentile, 75 percentile, and 25 percentile top and 25 percentile bottom. In addition, the continuous signature was applied to assess the association between signature and clinical outcome OS.    

### Median threshold

```{r signature analysis median, echo=TRUE}

# signature score
signature_score <- lapply(1:length(signature), function(k){ 
  
   sub_dat_expr <- dat_expr[rownames(dat_expr) %in% signature[[k]], ]
   apply(sub_dat_expr, 2, mean)
  
})

signature_score <- do.call(rbind, signature_score )
rownames(signature_score) <- names(signature)
colnames(signature_score) <- colnames(dat_expr)

# get low vs high signature score: median as threshold
signature_median <- lapply(1:nrow(signature_score), function(k){
    
    ifelse( as.numeric(as.character(signature_score[k,])) >= median(as.numeric(as.character(signature_score[k,]))) , "High" , "Low" )
    
}) 
  
signature_median  <- do.call(rbind, signature_median)
rownames(signature_median) <- rownames(signature_score)
colnames(signature_median) <- colnames(dat_expr)

# number of low vs high per signature

median_low_high_res <- lapply(1:nrow(signature_median), function(k){
     
     data.frame(signature_name = rownames(signature_median)[k],
                n_low = length(signature_median[k, ][signature_median[k,] == "Low"]),
                n_high = length(signature_median[k, ][signature_median[k,] == "High"]))
  
})

median_low_high_res  <- do.call(rbind, median_low_high_res)
write.csv(median_low_high_res, file="~/Result/LeucegeneAML/low_high_median.csv")

# KM plot along with log-rank test
  
for(j in 1:nrow(signature_median)){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_median[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     fit <- survfit(Surv(time, status) ~ sigScore, data = df)
     
      jpeg(file = paste(paste("~/Result/LeucegeneAML/ALL/KM/Median", 
                                    rownames(signature_median)[j],  sep="/"), ".jpg", sep=""),
     res=600, width=2700, height=2500)
    
      p <- ggsurvplot(fit, 
           data = df, 
           palette =
           c("Red", "Blue"),
           legend.labs = c("High", "Low"),
           pval = TRUE,
           pval.size = 3, 
           xlab = "Days",
           ylab = "Percent survival",
           title = paste("median", rownames(signature_median)[j], sep = ":"))
     print(p)
     dev.off()
    
  }
  

## get the cox results for meta-analysis
res_Cox <- lapply(1:nrow(signature_median), function(j){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_median[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     
      fit <- coxph(Surv(time, status) ~ sigScore, data = df) 
  
     data.frame(signature = rownames(signature_median)[j],
             n = summary(fit)$n,
             logHR = round(summary(fit)$coefficients[, "coef"], 3),
             SE = round(summary(fit)$coefficients[, "se(coef)"], 3),
             pval = summary(fit)$coefficients[, "Pr(>|z|)"]
             )
    
    
  })

res_Cox <- do.call(rbind, res_Cox)
res_Cox$FDR <- p.adjust(res_Cox$pval, method="BH")

write.csv(res_Cox, file="~/Result/LeucegeneAML/res_ALL_cox_median.csv")
print(res_Cox)

```


\newpage

### Percentile threshold (top 25\% and bottom 25\%)

```{r signature analysis percentile 25, echo=TRUE}

# get low vs high signature score: 75% Percentile as threshold
signature_percentile <- lapply(1:nrow(signature_score), function(j){
  
    qt <- quantile( as.numeric(as.character(signature_score[j,])) , na.rm=TRUE , probs= c( 0.25, 0.75 ) )
    ifelse( as.numeric(as.character(signature_score[j,])) <= qt[1] , "Low" , 
						ifelse( as.numeric(as.character(signature_score[j,])) <= qt[2] , "Median" , 
						ifelse( as.numeric(as.character(signature_score[j,])) > qt[2] , "High" , NA ) ) ) 
     }
   )

signature_percentile <- do.call(rbind, signature_percentile)
rownames(signature_percentile) <- names(signature)
colnames(signature_percentile) <- colnames(dat_expr)

# number of low vs high per signature
percentile_low_high_res <- lapply(1:nrow(signature_percentile), function(k){
     
     data.frame(signature_name = rownames(signature_percentile)[k],
                n_low = length(signature_percentile[k, ][signature_percentile[k,] == "Low"]),
                n_high = length(signature_percentile[k, ][signature_percentile[k,] == "High"]))
  
})

percentile_low_high_res  <- do.call(rbind, percentile_low_high_res)
write.csv(percentile_low_high_res, file="~/Result/LeucegeneAML/low_high_percentile_25_25.csv")
# KM plot along with log-rank test

for(j in 1:nrow(signature_percentile)){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_percentile[j, ])
     
     df <- df[df$sigScore != "Median", ]
     df <- df[!is.na(df$sigScore), ]
     fit <- survfit(Surv(time, status) ~ sigScore, data = df)
     
      jpeg(file = paste(paste("~/Result/LeucegeneAML/ALL/KM/Percentile_25", 
                                    rownames(signature_percentile)[j],  sep="/"), ".jpg", sep=""),
     res=600, width=2700, height=2500)
    
    
      p <- ggsurvplot(fit, 
           data = df, 
           palette =
           c("Red", "Blue"),
           legend.labs = c("High", "Low"),
           pval = TRUE,
           pval.size = 3, 
           xlab = "Days",
           ylab = "Percent survival",
           title = paste("Percentile", rownames(signature_percentile)[j], sep = ":"))
   print(p)
   dev.off()
    
  }
  
  
## get the cox results for meta-analysis
res_Cox <- lapply(1:nrow(signature_percentile), function(j){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_percentile[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     df <- df[df$sigScore != "Median", ]
     
      fit <- coxph(Surv(time, status) ~ sigScore, data = df) 
  
     data.frame(signature = rownames(signature_median)[j],
             n = summary(fit)$n,
             logHR = round(summary(fit)$coefficients[, "coef"], 3),
             SE = round(summary(fit)$coefficients[, "se(coef)"], 3),
             pval = summary(fit)$coefficients[, "Pr(>|z|)"]
             )
    
    
  })

res_Cox <- do.call(rbind, res_Cox)
res_Cox$FDR <- p.adjust(res_Cox$pval, method="BH")

write.csv(res_Cox, file="~/Result/LeucegeneAML/res_ALL_cox_percentile_25.csv")
print(res_Cox)

```


\newpage

### Percentile threshold 

```{r signature analysis percentile, echo=TRUE}

# get low vs high signature score: 75% Percentile as threshold
signature_percentile <- lapply(1:nrow(signature_score), function(j){
  
    qt <- quantile( as.numeric(as.character(signature_score[j,])) , na.rm=TRUE , probs= c( 0.75 ) )
  ifelse( as.numeric(as.character(signature_score[j,])) >= qt , "High" , "Low" )

  })

signature_percentile <- do.call(rbind, signature_percentile)
rownames(signature_percentile) <- names(signature)
colnames(signature_percentile) <- colnames(dat_expr)

# number of low vs high per signature
percentile_low_high_res <- lapply(1:nrow(signature_percentile), function(k){
     
     data.frame(signature_name = rownames(signature_percentile)[k],
                n_low = length(signature_percentile[k, ][signature_percentile[k,] == "Low"]),
                n_high = length(signature_percentile[k, ][signature_percentile[k,] == "High"]))
  
})

percentile_low_high_res  <- do.call(rbind, percentile_low_high_res)
write.csv(percentile_low_high_res, file="~/Result/LeucegeneAML/low_high_percentile.csv")
# KM plot along with log-rank test

for(j in 1:nrow(signature_percentile)){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_percentile[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     fit <- survfit(Surv(time, status) ~ sigScore, data = df)
     
      jpeg(file = paste(paste("~/Result/LeucegeneAML/ALL/KM/Percentile", 
                                    rownames(signature_percentile)[j],  sep="/"), ".jpg", sep=""),
     res=600, width=2700, height=2500)
    
    
      p <- ggsurvplot(fit, 
           data = df, 
           palette =
           c("Red", "Blue"),
           legend.labs = c("High", "Low"),
           pval = TRUE,
           pval.size = 3, 
           xlab = "Days",
           ylab = "Percent survival",
           title = paste("Percentile", rownames(signature_percentile)[j], sep = ":"))
   print(p)
   dev.off()
    
  }
  
  
## get the cox results for meta-analysis
res_Cox <- lapply(1:nrow(signature_percentile), function(j){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_percentile[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     
      fit <- coxph(Surv(time, status) ~ sigScore, data = df) 
  
     data.frame(signature = rownames(signature_median)[j],
             n = summary(fit)$n,
             logHR = round(summary(fit)$coefficients[, "coef"], 3),
             SE = round(summary(fit)$coefficients[, "se(coef)"], 3),
             pval = summary(fit)$coefficients[, "Pr(>|z|)"]
             )
    
    
  })

res_Cox <- do.call(rbind, res_Cox)
res_Cox$FDR <- p.adjust(res_Cox$pval, method="BH")

write.csv(res_Cox, file="~/Result/LeucegeneAML/res_ALL_cox_percentile.csv")
print(res_Cox)

```

\newpage

## Continuous signature

Consider the Cox proportional hazard model to assess the association between signature score (as continuous variable) and survival outcomes.  

```{r signature analysis continuous, echo=TRUE}

## get the cox results for meta-analysis
res_Cox <- lapply(1:nrow(signature_score), function(j){
    
     df <- data.frame(time = as.numeric(dat_clinical$`OS Days`),
                 status = as.numeric(dat_clinical$`OS Censoring Status`),
                 sigScore = signature_score[j, ])
     
     df <- df[!is.na(df$sigScore), ]
     
      fit <- coxph(Surv(time, status) ~ sigScore, data = df) 
  
     data.frame(signature = rownames(signature_median)[j],
             n = summary(fit)$n,
             logHR = round(summary(fit)$coefficients[, "coef"], 3),
             SE = round(summary(fit)$coefficients[, "se(coef)"], 3),
             pval = summary(fit)$coefficients[, "Pr(>|z|)"]
             )
    
    
  })

res_Cox <- do.call(rbind, res_Cox)
res_Cox$FDR <- p.adjust(res_Cox$pval, method="BH")

write.csv(res_Cox, file="~/Result/LeucegeneAML/res_ALL_cox_continuous.csv")

print(res_Cox)

```






